Bacterial Microbiome and Obesity Meta-analysis Supplemental

### 2016-03-18 microbiomeAndobesitySupplemental.Rmd
large supplemental updates

```{r data_table_loading, include=FALSE}
setwd("C:/Users/marc/Desktop/obesity2/HMP.analysis")

metaphlanData <- read.table("HMP_metaphlan2_merged_abundance_table.txt", header=T)
IDconversion <- read.csv("HMP.stool.IDs.csv")
microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)

#Only interested in first visit microbiome data
#need to create microbiome data set for only that

Data16S <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.

microb.rownames <- gsub("([0-9]+).*", "\\1", Data16S$Group) #extracting only numeric before first "."
rownames(Data16S) <- microb.rownames

#Get rid of information not used in downstram analysis
Data16S <- Data16S[, -c(1:3)]

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(Data16S)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- Data16S
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing, keep, microb.rownames, phyla.names)

#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
rm(phyla.total)

# Create comparable data set with 
metaphlanPhlyaData <- read.csv("HMP.metaphlan.phyla.data.csv", header=T)
rownames(metaphlanPhlyaData) <- metaphlanPhlyaData$ID
metaphlanPhlyaData <- metaphlanPhlyaData[,-1]
metaphlanPhlyaData <- as.data.frame(t(metaphlanPhlyaData))

#combine phyla that are not that abundant
metaphlanPhlyaData$other <- apply(metaphlanPhlyaData[, c("Acidobacteria", "Candidatus_Saccharibacteria", "Deinococcus_Thermus", "Fusobacteria", "Spirochaetes", "Synergistetes", "Tenericutes")], 1, sum)
metaphlanPhlyaData <- metaphlanPhlyaData[, -c(1, 4:5, 7, 9:11)]

#Create a relative abundance table for phyla
metaphlanPhlyaData.total <- apply(metaphlanPhlyaData[, c(1:6)], 1, sum)
metaphlanPhlyaData.rel.abund <- (metaphlanPhlyaData/metaphlanPhlyaData.total)*100

#Need to get sampleIDs to match 16S data
metaphlanNames <- rownames(metaphlanPhlyaData)
rownames(IDconversion) <- IDconversion[,1]
matchedData <- IDconversion[metaphlanNames, c(1,3)]
IndIDmatched <- matchedData[,2]

#Need to find and remove duplicated values (No idea which sample goes with which)
test1 <- as.data.frame(table(matchedData$Individual.ID))
good <- which(test1$Freq == 1)
unique <- test1[good, ]
uniqueNames <- as.character(unique$Var1)

test <- phyla.table.rel.abund[rownames(phyla.table.rel.abund) %in% uniqueNames, ]
finalUniqueNames <- rownames(test)
UpdatedMatchedData <- matchedData[matchedData$Individual.ID %in% finalUniqueNames, ]

goodMetaphlanPhylaData.rel.abund <- metaphlanPhlyaData.rel.abund[rownames(metaphlanPhlyaData.rel.abund) %in% rownames(UpdatedMatchedData), ]
rownames(goodMetaphlanPhylaData.rel.abund) <- UpdatedMatchedData[, 2]
testing <- goodMetaphlanPhylaData.rel.abund[order(rownames(goodMetaphlanPhylaData.rel.abund)), ]

phyla16SData <- test
phylaMetaphlanData <- testing
rm(finalUniqueNames, good, IndIDmatched, metaphlanNames, metaphlanPhlyaData.total, uniqueNames, goodMetaphlanPhylaData.rel.abund, matchedData, metaphlanPhlyaData, metaphlanPhlyaData.rel.abund, phyla.good, phyla.table, phyla.table.rel.abund, phylogenetic.info, test, test1, testing, unique, UpdatedMatchedData)

# Can start off with a general bar plot of the two side by side 
library(qdap)
library(plyr)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(gridExtra) #using grid.arrange() can add different plots together
library(plotrix)
library(ggExtra)
library(BlandAltmanLeh)

#phyla16SData <- phyla16SData[,-5] only if want to ignore unclassified
phylaMetaphlanData$unclassified <- 0
phylaMetaphlanData <- subset(phylaMetaphlanData, select=c(Actinobacteria:Proteobacteria, unclassified, Verrucomicrobia, other))
phyla16SData$dataset <- 1 
phylaMetaphlanData$dataset <- 2

```

**In-Depth Overview of Search Strategy:**

@@

From this total of 12 studies the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.  One study was excluded [@yatsunenko_human_2012] because it contained children and their sequencing of the 16S rRNA gene involved amplicons of only 100bp in length.  They also did not have obesity as part of their results in the actual published manuscript.  The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondance [@zeevi_personalized_2015].

This included a total of 10 studies in the qualitative synthesis of the analysis.  Because we decided a prioi to use the standard definition for BMI group classification one study from this ten did not have any individuals who were obese by this criteria [@nam_comparative_2011] and was excluded from the final quantitative synthesis and analysis.

@@

* PCR, qPCR, or RT-PCR used as main analysis Tool
* TRFLP or clone sequencing used to asses the bacterial community
* Utilization of 100bp or less single end reads for sequencing
* Sequencing Data not publicly available for download
* BMI not available and authors do not return correspondance
* Samples were not stool or feces
* Study contained children
* Study was a review

**A Comparison of 16S Amplicon Sequencing and Metaphlan2:**

```{r Figure_S1, echo=FALSE}
testset <- rbind(phyla16SData, phylaMetaphlanData)
testset$dataset <- factor(testset$dataset)
melt.testset <- melt(testset, id.vars=c("dataset"))
colnames(melt.testset)[2] <- "phyla"
barplotAnalysisData <- ddply(melt.testset, c("dataset", "phyla"), summarise, mean = mean(value), sd=sd(value), sem=sd(value/sqrt(length(value))))

bpg1 <- ggplot(barplotAnalysisData, aes(x=dataset, y=mean, fill=phyla)) + geom_bar(position = "fill", stat="identity")
bpg2 <- bpg1 + scale_fill_brewer(palette="Set1") + scale_y_continuous(label=percent)
bpg3 <- bpg2 + theme(axis.text=element_text(size=15), axis.title=element_text(size=20, face="bold"))
bpg4 <- bpg3 + xlab("Analysis Used") + ylab("Mean Relative Abundance (%)")
bpg5 <- bpg4 + guides(fill=guide_legend(title="Phyla")) + theme(legend.title = element_text(size=16, face="bold"))
bpg6 <- bpg5 + theme(axis.line=element_line(colour="black"), 
                 panel.grid.major = element_blank(), 
                 panel.grid.minor = element_blank(), 
                 panel.border = element_blank(), 
                 panel.background = element_blank())
bpg7 <- bpg6 + scale_x_discrete(breaks = 1:2, labels=c("16S", "Metaphlan2"))

bpg7
```

**Figure S1:Comparison of Phyla between 16S Sequencing and Metaphlan2.**

```{r Figure_S2, echo=FALSE, warning=FALSE, message=FALSE}
graphData <- as.data.frame(cbind(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes))
colnames(graphData) <- c("rRNA_16S", "Metaphlan2")

p1 <- ggplot(graphData, aes(x=rRNA_16S, y=Metaphlan2)) + geom_point(colour="blue", size=4)
p2 <- p1 + geom_smooth(method="lm", colour="black", se=TRUE)
p3 <- p2 + theme(axis.text=element_text(size=15), axis.title=element_text(size=20, face="bold"))
p4 <- p3 + xlim(0, 100) + xlab("16S") + ylim(0, 100)
p5 <- p4 + theme(axis.line=element_line(colour="black"), 
                 panel.grid.major = element_blank(), 
                 panel.grid.minor = element_blank(), 
                 panel.border = element_blank(), 
                 panel.background = element_blank())

ggExtra::ggMarginal(p5, type = "histogram")

```

**Figure S2: Bacteroidetes Linear Correlation between 16S Sequencing and Metaphlan2.**  Bacteroidetes has the best correlation of all the phyla (P-value = 0.0002277, R^2^ = 0.2358). The histogram for the 16S samples are on the x-axis while the histogram for Metaphlan2 is on the y-axis.

```{r Figure_S3, echo=FALSE}

ba <- bland.altman.plot(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes, graph.sys="ggplot2", main="Bland Altman Plot of Bacteroidetes", xlab="Means", ylab="Differences")

ba1 <- ba + theme(axis.text=element_text(size=15), axis.title=element_text(size=25, face="bold"))
ba2 <- ba1 + ylab("Difference") + xlab("Mean of Measurements")
ba3 <- ba2 + geom_point(size=4)
ba4 <- ba3 + theme(axis.line=element_line(colour="black"), 
                     panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(), 
                     panel.border = element_blank(), 
                     panel.background = element_blank())

ba4
```

**Figure S3: Bland-Altman plot of Bacteroidetes measurements for Metaphlan2 and 16S sequencing.**

>>>>

###In-Depth Overview of Search Strategy:

@@

From this total of 12 studies the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.  One study was excluded [@yatsunenko_human_2012] because it contained children and their sequencing of the 16S rRNA gene involved amplicons of only 100bp in length.  They also did not have obesity as part of their results in the actual published manuscript.  The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondance [@zeevi_personalized_2015].  A thrid study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]

Once these 3 studies were excluded there was a total of 9 studies in the qualitative synthesis of the analysis.  Because we decided a prioi to use the standard definition for BMI group classification one study from this ten did not have any individuals who were obese by this criteria [@nam_comparative_2011] and was excluded from the final quantitative synthesis and analysis.

@@

* PCR, qPCR, metagenomic sequencing, or RT-PCR used as main analysis Tool
* TRFLP or clone sequencing used to asses the bacterial community
* Utilization of 100bp or less single end reads for sequencing
* Sequencing Data not publicly available for download
* BMI not available and authors do not return correspondance
* Samples were not stool or feces
* Study contained children
* Study was a review

###Supplemental Results:

![Figure S1](results/figures/suppFigClostridiales.png)\

**Figure S1: Boxplots of All Clostridiales OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Clostridiales.  However, overall there was a wide range in direction both between studies and within studies.

![Figure S2](results/figures/suppFigLachno.png)\

**Figure S2: Boxplots of All Lachnospiraceae OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Lachnospiraceae.  However, overall there was a wide range in direction both between studies and within studies.

![Figure S3](results/figures/suppFigRumi.png)\

**Figure S3: Boxplots of All Ruminococcaceae OTUs Important for Non-Obese and Obese Classification.** A total of 7/8 studies had at least 1 OTU that was classified to Ruminococcaceae  However, overall there was a wide range in direction both between studies and within studies.

**Table 1. Summary of BLAST alignment of Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae classification OTUs**

```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}

library(knitr)

Study <- c("Wu", "Turnbaugh", "Ross", "Escobar", "Zupancic", "Baxter", "Goodrich", "HMP")
VariableRegion <- c("V1-V2", "V2", "V1-V3", "V1-V3", "V1-V3", "V4", "V4", "V3-V5")
ClosMostPrev <- c("OTU191:3 Clostridium", "None", "OTU102:8 Ruminococcus", "None", "OTU322:3 Parvimonas | OTU62:3 Parvimonas | OTU256:4 Clostridium", "OTU479:4 Anaerostipes", "OTU287:5 Clostridium | OTU662:8 Clostridium | OTU145:6 Clostridium | OTU244:3 Clostridium | OTU442:4 Clostridium", "OTU293:7 Clostridium") 
ClostHighMatch <- c("Anaerovorax odorimutans(81|95)", "None", "Ruminococcus champanellensis(100|100)", "None", "OTU322:Eubacterium hallii(92|100) | OTU62:Eubacterium hallii(92|99) | OTU256:Romboutsia lituseburensis(92|99)", "Anaerosporbacter mobilis(100|94)", "OTU287:Clostridium fimetarium(100|94) |  OTU662:Ruminiclostridium thermocellum(100|91) | OTU145:Ruminococcus lactaris(100|100) | OTU244:Anaerosporobacter mobilis(100|94) | OTU442:Clostridium propionicum(100|92)", "Anaerobacterium chartisolvens (100|91)")
LachMostPrev <- c("OTU75:6 Clostridium | OTU229:5 Peptostreptococcus | OTU5: 4 Blautia", "OTU318:", "OTU177:3 Roseburia", "OTU67:6 Eubacterium | OTU74:5 Blautia | OTU80:4 Blautia", "None", "OTU36:8 Clostridium", "OTU180:4 Clostridium | OTU75: 5 Clostridium | OTU122:5 Roseburia | OTU92:8 Clostridium | OTU673:7 Clostridium", "None")
LachHighMatch <- c("Clostridium bolteae(100|96)", "", "Clostridium Spp(100|98)", "OTU67:Dorea formicigenerans(100|96) | OTU74:Blautia luti(100|96) | OTU80:Blautia producta(100|93)", "None", "Lachnobacterium bovis(100|96)", "OTU180:Hungatella effluvii(100|95) | OTU75:Ruminococcus gnavus(100|100) | OTU122:Eubacterium ramulus(100|96) | OTU92:Fusicatenibacter saccharivorans(100|94) | OTU673:Bacteroides xylanolyticus(92|97) | OTU523:Eubacterium rectale(100|97)", "None")
RumiMostPrev <- c("OTU123:4 Clostridium", "OTU30:", "OTU22:3 Oscillibacter", "None", "OTU398:3 Eubacterium | OTU850:3 Ethanoligenens | OTU817:8 Ruminococcus", "OTU320:4 Clostridium", "OTU68:5 Ruminococcus | OTU255:3 Blautia | OTU1145:3 Ethanoligenens | OTU558:8 Ruminococcus | OTU297:3 Oscillibacter | OTU6:4 Clostridium | OTU132:2 Ruminoclostridium", "OTU252:10 Clostridium")
RumiHighMatch <- c("Anaerotruncus colihominis(85|95)", "", "Oscillibacter Spp(100|97)", "None", "OTU398:Oscillibacter valericigene(77|96) | OTU850:Papillibacter cinnamivorans(90|91) | OTU817:Ruminococcus callidus(91|95)", "Sporobacter termitidis(100|94)", "OTU68:Acetanaerobacterium elongatum(100|89) | OTU255:Pseudoflaconifractor capillosus(100|91) | OTU1145:Subdoligranulum variable(100|95) | OTU558:Ruminococcus champanellensis(92|98) | OTU297:Flavonifractor plautii(100|95) | OTU6:Oscillibacter valericigenes(100|93) | OTU132:Intestinimonas butyriciproducens(100|92)", "Clostridium Spp(100|97)")

table1 <- cbind(Study, VariableRegion, ClosMostPrev, ClostHighMatch, LachMostPrev, LachHighMatch, RumiMostPrev, RumiHighMatch)
colnames(table1) <- c("Study", "Hypervariable Region", "Clostridiales OTU Most Prevalent Match", "Clostridiales OTU Highest Match (Query Cover|Identity)", "Lachnospiraceae OTU Most Prevalent Match", "Lachnospiraceae OTU Highest Match (Query Cover|Identity)", "Ruminococcaceae OTU Most Prevalent Match", "Ruminococcaceae OTU Highest Match (Query Cover|Identity)")

kable(table1, format = "markdown", align = 'c', padding = 2)
```
